Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-14 content/contribution/continuous-datalog.tex
Fixed defn of streams

A {\em stream} $S$ is an unbounded, ordered sequence of fact
instances. We indicate that a fact instance $\langle f, t\rangle$
appears in $S$ by writing $\langle f, t \rangle \in S$. We indicate
that (in a given stream) fact instance $f_i$ appears before fact
instance $f_j$ by writing $f_i \prec f_j$.

Streams may be (but are not necessarily) ordered by the instants in
their fact instances, with the instants non-decreasing in the stream,
such that for any two fact instances $\langle f_i, t_i \rangle$ and
$\langle f_j, t_j\rangle$ where $\langle f_i, t_i \rangle \prec
\langle f_j, t_j \rangle$, $t_i \leqslant t_j$; we refer to such
streams as {\em in-order streams}, and to streams whose fact instances
are not ordered by their instants as {\em out-of-order streams}.
\end{definition}

@@

A window of range ${x \in \mathbb{N}}$ over a stream $S$ at an instant
${i \in \mathbb{T}}$ is the \emph{set} of fact instances that occur in
$S$ with instances in the interval ${(i-x,i]}$, written
  ${W^{S,x}_{i}}$, and defined as follows:
\[ W^{S,x}_i = \{ \langle f, t \rangle \mid \langle f, t \rangle \in S \land (i-x) < t \leqslant i \}\]  
\end{definition}

@@

However, this definition of a C-Datalog program does not take into
account any notion of windowing over the stream $S$. We define a {\em
  Windowed Continuous Datalog program} $\Pi^{C, r} = \langle R, F^0, S,
r \rangle$ to be a tuple with components as defined above for a
C-Datalog program, but with the addition of $r \in \mathbb{N}$ for the
range of the window over $S$, which effectively specifies that the
fact instances from $S$ are to contribute to the entailments of $\Pi^{WC}$
from their instant for a duration of $r$.

We consider the entailment of the WC-Datalog program $\Pi^{C,r}$ not
as a single stream of fact instances (as for C-Datalog), but in three
parts: a set of fact instances $E^0 = cons^T(\langle R, F^0\rangle$
that is the entailment of the TA-Datalog program $\langle R,
F^0\rangle$, a stream $P$ of fact instances entailed by the fact
instances in $F^0$ and $S$, and a stream $N$ of fact instance
negations: $\langle R, F^0, S, r \rangle \models \langle E^0, P, N
\rangle$. $P$ and $N$ can be considered to be the continuous reasoning
equivalents of the streams produced by the \emph{IStream} and
\emph{DStream} operators respectively of the \emph{CQL} continuous
query language \citep{CQL}, which create streams based on the
additions to and deletions from a relation. Note that $P$ is an
in-order stream whose members are a subset of those in $E^S$ for an
equivalent C-Datalog programe, while $N$ may be an out-of-order
stream.

%%%FIXME
In order to

%%%FIXME
From this description of instance entailment,
Axiom~\ref{axiom:continuous datalog: entailment disjointness} may be
derived:

\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}
  If two sliding windows over the same stream $S$ cover non-overlapping time periods,
then a set $E^T_1$ of fact instances entailed by ${\langle R, W^{S,x}_i \rangle}$
is necessarily disjoint from any set $E^T_2$ of fact instances entailed by
$\langle R, W^{S,y}_{i-z} \rangle$, where ${z \geq x}$.
% This is because the timestamps of the instances ${E^T1}$ are a subset of those of
% the instances $W^{S,x}_i$, the timestamps of the instances ${E^T2}$ are a subset
% of those of the instances $W^{S,y}_{i-z}$, fact instances are distinct by timestamp and
% ${\{ t : f | t \in W^{S,x}_i\} \cap \{ t : f | t \in W^{S,y}_{i-z}\} = \varnothing}$.
\[
\{ e \mid \langle R , W^{S,x}_i \rangle \models e \} \cap \{ e \mid \langle R , W^{S,y}_{i-z} \rangle \models e \} = \varnothing
\]
\end{axiom}

@@

The instantaneous entailments of a $\Pi^{C,r}_i$ are the union of the set
of persistent entailments of ${\langle R, F^0 \rangle \vDash E^0}$
(which are guaranteed to be entailed by every state, as positive
Datalog is monotonic), and the set $E^T_{i}$ of transient entailments
whose justifications are at least partially contained in
$W^{S,r}_i$. We define $E^T_i$ as:
\[
E^T_i = \{ x \mid \Pi^{WC}_i \models x \land J^T \cup W^{S,r}_i \neq \varnothing \}
\]
\noindent where $J^T$ is a justification for the entailment of x by $\Pi^{WC}_i$.

By this definition and Axiom~\ref{axiom:continuous datalog: entailment disjointness},
$E^0$ is guaranteed to be disjoint from every $E^T_i$
for every state $\Pi^{C,r}_i$ of every WC-Datalog program $\Pi^{C,r}$.
\end{definition}

@@

From this, we can see that each entailed fact instance $\langle f, e
\rangle$ in some $E^T_{i}$ has an instant $e$ that is associated with
the state $\Pi^{C,r}_e$ in which a justification for $\langle f, e
\rangle$ is first completed.

As such, the stream of entailed fact instances $P$ of
a $\Pi^{C,r}$ takes precisely the same form as the streams of facts already
described, thus producing a concise stream of entailments that may be
read by further stream processing.  In addition, the stream of negations
$N$ of a $\Pi^C$ also takes this form, but differs from the streams
discussed thus far in that it is not ordered by the entailment/assertion
time of the entailed fact instances, but by the real-time associated
with the first state $\Pi^C_{i_{n}}$ in which their justification
\emph{ceases} to be complete, i.e. ${just(\Pi^C_{i_{e}}, f|t) \subseteq W^{S,r}_{i_n - 1}}$
but ${just(\Pi^C_{i_{e}}, f|t) \not\subseteq W^{S,r}_{i_n}}$.

>>>>

A {\em stream} $S$ is an unbounded sequence of fact instances. We
define a sequence as a mapping from fact instances to instants, where
the instant records the time at which the fact instance appears on the
stream. This has the consequence that a particular fact instance
(i.e. a fact accompanied by a particular instant that is its time
annotation) may appear only once in a given stream. For convenience,
we indicate that a fact instance $\langle f, t\rangle$ appears in $S$
by writing $\langle f, t \rangle \in S$. We indicate that a fact
instance $f_i$ appears strictly before a fact instance $f_j$ on stream
$S$ (i.e. $S(f_i) < S(f_j)$) by writing $f_i \prec f_j$.

Streams may be (but are not necessarily) ordered by the instants in
their fact instances, with the instants non-decreasing in the stream,
such that for any two fact instances $\langle f_i, t_i \rangle$ and
$\langle f_j, t_j\rangle$ where $\langle f_i, t_i \rangle \prec
\langle f_j, t_j \rangle$, $t_i < t_j$; we refer to such
streams as {\em in-order streams}, and to streams whose fact instances
are not ordered by their instants as {\em out-of-order streams}.
\end{definition}

@@

A window of range ${r \in \mathbb{N}}$ over a stream $S$ at an instant
${i \in \mathbb{T}}$ is the \emph{set} of fact instances that appear in
$S$ in the interval ${(i-r,i]}$, written ${W^{S,r}_{i}}$, and defined as follows:
\[
W^{S,r}_i = \{ \langle f, t \rangle \mid \langle f, t \rangle \in S \land (i - r) < S(\langle f, t\rangle) \leqslant i \}
\]  
\end{definition}

@@

However, this definition of a C-Datalog program does not take into
account any notion of windowing over the stream $S$. We define a {\em
  Windowed Continuous Datalog program} $\Pi^{C, r} = \langle R, F^0, S,
r \rangle$ to be a tuple with components as defined above for a
C-Datalog program, but with the addition of $r \in \mathbb{N}$ for the
range of the window over $S$, which effectively specifies that the
fact instances from $S$ are to contribute to the entailments of $\Pi^{C,r}$
from their instant for a duration of $r$.

We consider the entailment of the WC-Datalog program $\Pi^{C,r}$ not
as a single stream of fact instances (as for C-Datalog), but in three
parts: a set of fact instances $E^0 = cons^T(\langle R, F^0\rangle)$
that is the entailment of the TA-Datalog program $\langle R,
F^0\rangle$, a stream $P$ of fact instances entailed by the fact
instances in $F^0$ and $S$, and a stream $N$ of fact instance
negations: $\langle R, F^0, S, r \rangle \models \langle E^0, P, N
\rangle$. $P$ and $N$ can be considered to be the continuous reasoning
equivalents of the streams produced by the \emph{IStream} and
\emph{DStream} operators respectively of the \emph{CQL} continuous
query language \citep{CQL}, which create streams based on the
additions to and deletions from a relation. Note that $P$ is an
in-order stream whose members are a subset of those in $E^S$ for an
equivalent C-Datalog programe, while $N$ may be an out-of-order
stream.
\end{definition}

From the definition of entailment for TA-Datalog in \refdef{continuous
  datalog: ta datalog program} and the definition of windows in
\refdef{continuous datalog: window}, we can derive the following:

\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}

If two sliding windows over the same in-order stream $S$ cover
non-overlapping time periods, then the set of fact instances entailed
by ${\langle R, W^{S,x}_i \rangle}$ is necessarily disjoint from the
set of fact instances entailed by $\langle R, W^{S,y}_{i-z} \rangle$,
where ${z \geq x}$.
\[
\{ e \mid \langle R , W^{S,x}_i \rangle \models e \} \cap \{ e \mid \langle R , W^{S,y}_{i-z} \rangle \models e \} = \varnothing
\]
\end{axiom}

@@

The instantaneous entailments of a $\Pi^{C,r}_i$ are the union of the set
of persistent entailments of ${\langle R, F^0 \rangle \vDash E^0}$
(which are guaranteed to be entailed by every state, as positive
Datalog is monotonic), and the set $E^T_{i}$ of transient entailments
whose justifications are at least partially contained in
$W^{S,r}_i$. We define $E^T_i$ as:
\[
E^T_i = \{ x \mid \Pi^{C,r}_i \models x \land J^T \cup W^{S,r}_i \neq \varnothing \}
\]

\noindent where $J^T$ is a justification for the entailment of x by
$\Pi^{C,r}_i$. By this definition and Axiom~\ref{axiom:continuous
  datalog: entailment disjointness}, $E^0$ is guaranteed to be
disjoint from every $E^T_i$ for every state $\Pi^{C,r}_i$ of every
WC-Datalog program $\Pi^{C,r}$.
\end{definition}

@@

As noted above in \refdef{continuous datalog: CDP}, the stream of
negations $N$ of a WC-Datalog program $\Pi^{C,r}$ may be an
out-of-order stream in which fact instances appear on the stream

also takes this form, but differs from the streams
discussed thus far in that it is not ordered by the entailment/assertion
time of the entailed fact instances, but by the real-time associated
with the first state $\Pi^C_{i_{n}}$ in which their justification
\emph{ceases} to be complete, i.e. ${just(\Pi^C_{i_{e}}, f|t) \subseteq W^{S,r}_{i_n - 1}}$
but ${just(\Pi^C_{i_{e}}, f|t) \not\subseteq W^{S,r}_{i_n}}$.
